COVID19 - Forecast analysis - time dependent SEIR model - Veneto

### 2020-03-14 draft_analysis_Italy.Rmd
Add home page, navigation and theme

The present analysis started from the dataset on COVID19 updated in https://github.com/pcm-dpc/COVID-19, database provided by the Italian Civil Protection.  
```{r setup}
rm(list=ls())
###import italian dataset updated 13 March 2020
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-andamento-nazionale/dpc-covid19-ita-andamento-nazionale.csv",header=T)
days<-dim(dat_csv)[1]
dat_csv$t<-1:days
#database contains  the following variables
names(dat_csv)
# The total number of epidemic day is
days

With the aim to predict the future number of COVID19 cases on the basis of the actual data, we used a SEIR model applied to the COVID19 epidemic in Italy.
A first parameter to estimate is R0, the reproduction number (https://en.wikipedia.org/wiki/Basic_reproduction_number).   R0 indicates how contagious an infectious disease is. It is also referred to as "the reproduction number" of COVID19.

@@

where $`Y_t`$ is the cumulative number of infected at the time t, while b is beta, the slope of the regression line.

@@

The R0 shows a decreasing trend in the last period. The slope b indicates the rate of exponential increase.  
However,  R0 is going to decrease in the next days. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days assuming a Gamma distribution for the beta (the slope) forcing its value to be greater than 0. The trend was not monotonic, we use a simple splines to increase the fitting of the model to the data.
```{r R0 forecast}
library(splines)
time<-1:(days-4)
# splines with only 3 basis, weighted for standard deviation of estimates.
beta.model<-glm(beta_vec~bs(time,3),weights = 1/sd_vec,family=Gamma)
forecast=14
# add 'fit', 'lwr', and 'upr' columns to dataframe (generated by predict)
pre<-predict(beta.model,type='response',newdata=data.frame(time=1:(days+forecast)),se.fit=TRUE)
date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
beta.predict <- data.frame(beta_vec=c(beta_vec,rep(NA,forecast+4)),time=date,fit=pre$fit,lwr=pre$fit-1*1.96*pre$se.fit,upr=pre$fit+1*1.96*pre$se.fit)

@@

We want to make a short term forecast (14 days) with 3 scenario:

-Scenario 1: 10 exposed people for each COVID-19 case (no restrictions made or even no effects)

@@

We made a forecast by means of a SEIR model fixing a series of initial status:  
- S0=N, Italian population   
- E= f * I0 (with f a fixed factor of the previous scenario)  
- I0: initial number of COVID-19 cases  
- R0: initial number of recovered

@@

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$I*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$I*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$I*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

@@

plot(date,c(dat_csv$totale_casi,(seir1$I+seir1$R)*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$I+seir2$R)*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$I+seir3$R)*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

>>>>

The present analysis started from the dataset on COVID19 updated in https://github.com/pcm-dpc/COVID-19, database provided by the Italian Civil Protection.  
```{r setup}
rm(list=ls())
###import italian dataset updated 13 March 2020
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-andamento-nazionale/dpc-covid19-ita-andamento-nazionale.csv",header=T)
days<-dim(dat_csv)[1]
dat_csv$t<-1:days
#database contains  the following variables
names(dat_csv)
# The total number of epidemic day is
days
plot(as.Date(substr(dat_csv$data,1,10)),dat_csv$totale_attualmente_positivi,ylab=" Infected",xlab="Date",type="l")
points(as.Date(substr(dat_csv$data,1,10)),dat_csv$totale_attualmente_positivi)
```

With the aim of predicting the future number of COVID19 cases on the basis of the actual data, we used a SEIR model applied to the COVID19 epidemic in Italy.
A first parameter to estimate is R0, the reproduction number (https://en.wikipedia.org/wiki/Basic_reproduction_number).   R0 indicates how contagious an infectious disease is. It is also referred to as "the reproduction number" of COVID19.

@@

where $`Y_t`$ is the  number of infected at the time t, while b is beta, the slope of the regression line.

@@

The R0 shows a decreasing trend in the last period. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days assuming a Gamma distribution for the beta (the slope) forcing its value to be greater than 0. The trend was not monotonic, we use a simple splines to increase the fitting of the model to the data.
```{r R0 forecast}
library(splines)
time<-1:(days-4)
# splines with only 3 basis, weighted for standard deviation of estimates.
beta.model<-glm(beta_vec~bs(time,3),weights = 1/sd_vec,family=Gamma)
forecast=14
# add 'fit', 'lwr', and 'upr' columns to dataframe (generated by predict)
pre<-predict(beta.model,type='response',newdata=data.frame(time=1:(days+forecast)),se.fit=TRUE)
date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
beta.predict <- data.frame(beta_vec=c(beta_vec,rep(NA,forecast+4)),time=date,fit=pre$fit,lwr=pre$fit-1*1.96*pre$se.fit,upr=pre$fit+1*1.96*pre$se.fit)

@@

We want to make a short term forecast (14 days) with 3 scenario, based on the number of exposed people:

-Scenario 1: 10 exposed people for each COVID-19 case (no home restrictions made or even no effects)

@@

We made a forecast by means of a SEIR model fixing a series of initial status:  
- S0=N, the size of Italian population   
- E= f * I0 (with f a fixed factor of the previous scenario)  
- I0: initial number of COVID-19 cases  
- R0: initial number of recovered

@@

date<-seq(as.Date("2020-02-24"),as.Date("2020-02-24")+forecast-1+dim(dat_csv)[1],1)
plot(date,c(dat_csv$totale_attualmente_positivi,seir1$I*N),type="l",ylab="Cases",xlab="time",main="Infected")
lines(date,c(dat_csv$totale_attualmente_positivi,seir2$I*N),col=2)
lines(date,c(dat_csv$totale_attualmente_positivi,seir3$I*N),col=3)
lines(date[1:dim(dat_csv)[1]],dat_csv$totale_attualmente_positivi,lwd=2)
legend("topleft",c("first scenario - Exp=10*I","second scenario Exp=5*I","third scenario Exp=3*I"),lty=1,col=1:3)
```

@@

plot(date,c(dat_csv$totale_casi,(seir1$I+seir1$R)*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$I+seir2$R)*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$I+seir3$R)*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("Scenario 1 -Exp=10*I","Scenario 2 - Exp=5*I","Scenario 3- Exp=3*I"),lty=1,col=1:3)
```

### 2020-03-14 draft_analysis_Veneto.Rmd
Add home page, navigation and theme

The present analysis used the dataset on COVID19 updated in https://github.com/pcm-dpc/COVID-19.  
We used a SEIR model to predict the trend of the actual COVID19 epidemic in the Veneto Region.

```{r setup, include=TRUE}
rm(list=ls())
###import italian dataset updated 13 March 2020 - 
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv",header=T)
#we restricted the analysis to the Veneto Region
dat_csv<-dat_csv[dat_csv$codice_regione==5,]
dat_csv$t<-1:dim(dat_csv)[1]
plot(dat_csv$data,dat_csv$totale_attualmente_positivi,ylab="Infected",xlab="Date")
days<-dim(dat_csv)[1]
```

The actual status is

@@

The R0 shows a decreasing trend in the last period. The slope b indicates the rate of exponential increase.  
However,  R0 is going to decrease in the next days. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days assuming a Gamma distribution for the beta (the slope) forcing its value to be greater than 0. The trend was not monotonic, we use a simple splines to increase the fitting of the model to the data.
```{r R0 forecast}
library(splines)
time<-1:(days-4)

@@

-Scenario 1: 10 exposed people for each COVID-19 case (no restrictions made or even no effects)

@@

We made a forecast by means of a SEIR model fixing a series of initial status:  
- S0=N, Italian population   
- E= f * I0 (with f a fixed factor of the previous scenario)  
- I0: initial number of COVID-19 cases  
- R0: initial number of recovered

@@

```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
# italian poulation
N=4905854
# duration of COVID19 
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics) and the function SEIR() to implement a SEIR model:

@@

plot(date,c(dat_csv$totale_casi,(seir1$I+seir1$R)*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$I+seir2$R)*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$I+seir3$R)*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("first scenario","second scenario","third scenario"),lty=1,col=1:3)
```

>>>>

The present analysis is based on the COVID19 dataset updated in https://github.com/pcm-dpc/COVID-19.  
This dataset contained the time series of infected, recovered, deaths, etc...for each day after 24/02/2020.
We used a SEIR model to predict the trend of the actual COVID19 epidemic in the Veneto Region.

```{r setup, include=TRUE}
rm(list=ls())
###import italian dataset updated 13 March 2020 - 
dat_csv<-read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv",header=T)
#we restricted the analysis to the Veneto Region
dat_csv<-dat_csv[dat_csv$codice_regione==5,]
dat_csv$t<-1:dim(dat_csv)[1]
plot(as.Date(substr(dat_csv$data,1,10)),dat_csv$totale_attualmente_positivi,ylab=" Infected",xlab="Date",type="l")
points(as.Date(substr(dat_csv$data,1,10)),dat_csv$totale_attualmente_positivi)
days<-dim(dat_csv)[1]
```

The actual time series are

@@

The R0 shows a decreasing trend in the last period. We use the estimated trend between R0 and time to calculate the future R0 value for the next 14 days. 
We predict beta (and R0) for the next 14 days assuming a Gamma distribution for the beta (the slope) forcing its value to be greater than 0. The trend was not monotonic, we use a simple splines to increase the fitting of the model to the data.
```{r R0 forecast}
library(splines)
time<-1:(days-4)

@@

-Scenario 1: 10 exposed people for each COVID-19 case (no home restrictions made or even no effects)

@@

We made a forecast by means of a SEIR model fixing a series of initial status:  
- S0=N, the size of Veneto Region population 
- E= f * I0 (with f a fixed factor of the previous scenario)  
- I0: initial number of COVID-19 cases  
- R0: initial number of recovered

@@

```{r initation plot, echo=TRUE}
# initial number of infectus
I0<-dat_csv$totale_attualmente_positivi[dim(dat_csv)[1]]; I0
# initial number of recovered
R0<-dat_csv$dimessi_guariti[dim(dat_csv)[1]]; R0
# Veneto Region population
N=4905854
# duration of COVID19 diseases before get recovered
duration<-14
#sigma0 is the coronavirus transmission rate fixed to 5%  (half of flu epidemic)
sigma0<-0.05
#mortality rate 
mu0<-1/(82*365.25) # 1/lifespan
```
We use the library(EpiDynamics) and the function SEIR() to implement a SEIR model:

@@

plot(date,c(dat_csv$totale_casi,(seir1$I+seir1$R)*N),type="l",ylab="Cases",xlab="time",main="Cumulative Infected")
lines(date,c(dat_csv$totale_casi,(seir2$I+seir2$R)*N),col=2)
lines(date,c(dat_csv$totale_casi,(seir3$I+seir3$R)*N),col=3)
lines(date[1:dim(dat_csv)[1]],(dat_csv$totale_casi),lwd=2)
legend("topleft",c("Scenario 1 -Exp=10*I","Scenario 2- Exp=5*I","Scenario 3- Exp=3*I"),lty=1,col=1:3)
```
